Uncertainty propagation in deep neural networks for sparse coding

### 2018-03-09 paper/BayesianLISTA.tex
small updates to text

\section{Introduction}
Though the idea of Bayesian learning in neural networks is not new \citep{neal2012bayesian}, it has regained its attention recently with the development of distributed approximate inference techniques \citep{li2015stochastic, hoffman2013stochastic}  and general boost in popularity of deep learning. In some spheres, such as self-driving cars or healthcare, uncertainty estimates are very important and Bayesian learning allows to obtain them.

@@

\section{Neural networks for sparse coding}
\label{sec:nn_sc}
The sparse coding and sparse inverse problems are the linear regression with the assumption of sparse weights
\begin{equation}
\label{eq:regression_problem}
\mathbf{y} = \mathbf{X}\boldsymbol\beta + \varepsilon
\end{equation}
where $\mathbf{y} \in \mathbb{R}^K$ is the observation, $\mathbf{X} \in \mathbb{R}^{K \times D}$ is the design matrix, $\boldsymbol\beta \in \mathbb{R}^D$ is the unknown vector of weight with many elements $\beta_d$ equal to zero. $\varepsilon \in \mathbb{R}^K$ is the independent Gaussian noise.

@@

\section{Bayesian neural network for sparse coding}
\label{sec:bayesian_lista}

This section presents the proposed Bayesian neural network for sparse coding. To formulate the Bayesian version of \textsc{lista} we impose the prior distributions on the unknown weights
\begin{align}
&p(\mathbf{W}) = \prod_{d=1}^D\prod_{k=1}^K \mathcal{N}(w_{ij} | 0, \eta^{-1}), \\
&p(\mathbf{S}) = \prod_{d'=1}^D\prod_{d''=1}^D \mathcal{N}(s_{d'd''} | 0, \eta^{-1}),
\end{align}
where $w_{ij}$ is a component of the matrix $\mathbf{W}$, $s_{d'd''}$ is a component of the matrix $\mathbf{S}$, $\eta$ is the precision of the Gaussian distribution.
The likelihood of $\mathbf{B}$ is defined as 
\begin{align}
p(&\mathbf{B}| \mathbf{Y}, \mathbf{W}, \mathbf{S}, \gamma, \lambda) \nonumber \\
\label{eq:likelihood}
= &\prod_{n=1}^N\prod_{d=1}^D\mathcal{N}(\beta_d^{(n)}; [f(\mathbf{y^{(n)}}; \mathbf{W}, \mathbf{S}, \lambda)]_d, \gamma^{-1}),
\end{align}
where $[\cdot]_d$ is the d-th component of a vector, $f(\mathbf{y}^{(n)}; \mathbf{W}, \mathbf{S}, \lambda)$ is the output of the Bayesian \textsc{lista} neural network, $\gamma$ is the precision of the Gaussian distribution.
The prior of the introduced Gaussian precisions are set to the Gamma distribution with parameters $a^{\cdot}$ and $b^{\cdot}$:
\begin{align}
&p(\gamma) = \text{Gam}\left(\gamma; a^{\gamma}, b^{\gamma}\right), \\
&p(\eta) = \text{Gam}\left(\eta; 	a^{\eta}, b^{\eta}\right)
\end{align}

@@

\section{Uncertainty propagation through soft thresholding}
\label{sec:fprop}
For every layer of the Baysian \textsc{lista} we assume that $\widehat{\boldsymbol\beta}_{l-1}$ has the spike and slab distribution with parameters: $\boldsymbol\omega$ is a probability of a spike, $\mathbf{m}$ is a mean of a slab Gaussian distribution, and $\mathbf{v}$ is a variance of the slab distribution:
\begin{equation}
[\boldsymbol\beta_{l-1}]_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d),
\end{equation}
where $\omega_d$, $m_d$, and $v_d$ are components of vectors $\boldsymbol\omega$, $\mathbf{m}$, and $\mathbf{v}$, respectively, $\delta_0$ is the delta-function that represents a spike. 
We initialise $\widehat{\boldsymbol\beta}_{0}$ with the spike and slab distribution with parameters $\boldsymbol\omega = \mathbf{0}$, $\mathbf{m} = \mathbf{0}$, and $\mathbf{v} = \mathbf{1}$.

@@

  \begin{lemma}[Product of Gaussian matrix and spike and slab vector]
  \label{thm:matrix_vector}
 Let $\mathbf{S} \in \mathbb{R}^{D \times D}$ be a matrix of independent Gaussian-distributed random variables: $s_{d'd''} \sim \mathcal{N}(m^s_{d'd''}, v^s_{d'd''})$, and $\boldsymbol\beta \in \mathbb{R}^D$ be a vector with spike-and-slab distributed variables: $\beta_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d)$. If their product is approximated as a vector $\mathbf{e} \in \mathbb{R}^{D}$ of random variables $e_d \sim \mathcal{N}(m^e_d, w^e_d)$, then 
\begin{subequations}
\begin{align}
 m^e_d &= \sum_{d'=1}^D m^s_{dd'}(1-\omega_{d'})m_{d'}, \\
 \begin{split}
 v^e_d &= \sum_{d'=1}^D [(m^s_{dd'})^2(1-\omega_{d'})^2v_{d'} \\
 & {}+ (1-\omega_{d'})^2(m_{d'})^2v^s_{dd'} + v^s_{dd'}(1-\omega_{d'})^2v_{d'}].
 \end{split}
 \end{align}
\end{subequations}
 \end{lemma}
 \begin{proof}
 	We compute the mean and variance of the product $\mathbf{S}\boldsymbol\beta$ with an assumption of independence of its components and approximate resulting distribution with Gaussian. The quality of this approximation is discussed in section \ref{sec:approx_quality}
\begin{flalign*}
	\mathbb{E}e_d &= \sum_{d'=1}^D \mathbb{E}[s_{dd'}\beta_{d'}]  = \sum_{d'=1}^D m^s_{dd'}\mathbb{E}\beta_{d'}\\
	\operatorname{Var}e_d &= \sum_{d'=1}^D \operatorname{Var}[s_{dd'}\beta_{d'}] = \sum_{d'=1}^D [(\mathbb{E}s_{dd'})^2 \operatorname{Var}\beta_{d'} \\
	&{}+ (\mathbb{E}\beta_{d'})^2 \operatorname{Var}s_{dd'} + \operatorname{Var}\beta_{d'} \operatorname{Var}s_{dd'}]
\end{flalign*}
where $\mathbb{E}\beta_{d'}$, $\operatorname{Var}\beta_{d'}$ are computed according to lemma~\ref{thm:moments_spsl}.
 \end{proof}

@@

The proposed operations on Gaussian and spike-and-slab distributions allow to formulate the main result in this section that describes how the uncertainty propagation works.
\begin{theorem}[Bayesian \textsc{lista} forward propagation]
\label{thm:prob_layer}
The spike and slab distribution for $\boldsymbol\beta$ with can be propagated through the \textsc{lista} layer and the parameters after propagation can be computed as following:
\begin{enumerate}
	\item $\mathbf{b} = \mathbf{W}\mathbf{y}$ is computed according to lemma \ref{thm:matrix_const};
	\item $\mathbf{e} = \mathbf{S}\boldsymbol\beta_{l-1}$ is computed according to lemma \ref{thm:matrix_vector};
	\item $\mathbf{c} = \mathbf{b} + \mathbf{e}$ is computed according to lemma \ref{thm:sum_vectors};
	\item $\boldsymbol\beta_{l} = h_\lambda(\mathbf{c})$ is computed according to lemma \ref{thm:soft_thresholding}.
\end{enumerate}
\end{theorem}

\subsection{Approximation quality}
\label{sec:approx_quality}
We have used two approximations in forward propagation of uncertainty. First, in lemma \ref{thm:matrix_vector} a Gaussian matrix is multiplied by a spike-and-slab vector and their product is approximated with the Gaussian distribution. Second, in lemma \ref{thm:soft_thresholding} the result of soft thresholding of a Gaussian vector is approximated with the spike-and-slab distribution. In this section we demonstrate that these approximations are close to the real distributions.

@@

\section{Backpropagation}
\label{sec:backpropagation}
\subsection{Likelihood}
For Bayesian inference we significantly expend the probabilistic backpropagation algorithm~\citep{hernandez2015probabilistic} for computing parameter updates. It is based on assumed density filtering (ADF) and expectation propagation (EP) and allows to update parameters of the distributions based on the derivative of the logarithm of a normalisation constant. ADF is based on iteratively incorporating factors from the true posterior to a factorised approximating distribution, whereas in EP factors in an approximating distribution are iteratively replaced by factors from the true posterior.

The exact posterior (\ref{eq:posterior}) is approximated with a factorised distribution
\begin{align}
\begin{split}
q(\mathbf{W}, \mathbf{S}, \gamma, \eta) &= \prod_{d=1}^D\prod_{k=1}^K \mathcal{N}(w_{dk} | m^w_{dk}, v^w_{dk}) \\
&\times \prod_{d'=1}^D\prod_{d''=1}^D \mathcal{N}(s_{d'd''} | m^s_{d'd''}, v^s_{d'd''}) \\
&\times \text{Gam}(\gamma; a^\gamma, b^\gamma) \text{Gam}(\eta; a^\eta, b^\eta) 
\end{split}
\end{align}

With the likelihood factors~(\ref{eq:likelihood}) of the true posterior we employ the ADF approach and iteratively incorporate them into the approximating distribution $q$. The normalisation constant of the approximating distribution $q$ with the likelihood term for the data point $n$ incorporated can be computed as follows (to simplify notation the superscript $(n)$ is omitted)
\begin{align}
\label{eq:Z}
\begin{split}
Z = \int \prod_{d=1}^{D} \big[&\mathcal{N}(\beta_d | f(\mathbf{y} ; \mathbf{S}, \mathbf{W}, \lambda), \gamma^{-1}) \\
 &q(\mathbf{W}, \mathbf{S}, \gamma, \eta)\big] \mathrm{d}\mathbf{W} \mathrm{d}\mathbf{S} \mathrm{d}\gamma \mathrm{d}\eta
 \end{split}
\end{align}
We sample $\mathbf{W}$, $\mathbf{S}$ from $q$ and get $\mathbf{z}_L = f(\mathbf{y} ; \mathbf{S}, \mathbf{W}, \lambda)$, that is the output from the network and approximated with the spike and slab distribution with parameters $\boldsymbol\omega^{z_L}$, $\mathbf{m}^{z_l}$, and $\mathbf{v}^{z_L}$
\begin{align}
&Z \approx \int \prod_{d=1}^{D} \big[\mathcal{N}(\beta_d | [\mathbf{z}_L]_d, \gamma^{-1}) \nonumber \\
&\times (\omega^{z_L}_d \delta_0([\mathbf{z}_L]_d) + (1 - \omega^{z_L}_d)\mathcal{N}([\mathbf{z}_L]_d | m^{z_L}_d, v^{z_L}_d)) \nonumber\\
&\times \text{Gam} (\gamma_d; \alpha^\gamma, \beta^\gamma)\big]\mathrm{d}\mathbf{z}_L \mathrm{d}\gamma  \nonumber\\
&= \prod_{d=1}^{D} \Big[\omega^{z_L}_d \int \big[\mathcal{N}(\beta_d | [\mathbf{z}_L]_d, \gamma^{-1})  \delta_0([\mathbf{z}_L]_d) \nonumber \\
&\times \text{Gam} (\gamma; \alpha^\gamma, \beta^\gamma)\big]\mathrm{d}[\mathbf{z}_L]_d \mathrm{d}\gamma  \nonumber\\
& + (1 - \omega^{z_L}_d)\int \big[\mathcal{N}(\beta_d | [\mathbf{z}_L]_d, \gamma^{-1})\nonumber\\
&\times \mathcal{N}([\mathbf{z}_L]_d | m^{z_L}_d, v^{z_L}_d)) \text{Gam} (\gamma; \alpha^\gamma, \beta^\gamma)\big]\mathrm{d}[\mathbf{z}_L]_d \mathrm{d}\gamma\Big]  \nonumber\\
& = \prod_{d=1}^{D} \Big[\omega^{z_L}_d \int \mathcal{N}(\beta_d | 0, \gamma^{-1})  \text{Gam} (\gamma; \alpha^\gamma, \beta^\gamma) d\gamma  \nonumber\\
& + (1 - \omega^{z_L}_d)\int \big[\mathcal{T}(\beta_d | [\mathbf{z}_L]_d, \frac{\beta^\gamma}{\alpha^\gamma}, 2\alpha^\gamma)  \nonumber\\
&\times \mathcal{N}([\mathbf{z_L}]_d| m^{z_L}_d, v^{z_L}_d))\big] \mathrm{d}[\mathbf{z}_L]_d\Big]  \nonumber\\
& \approx \prod_{d=1}^D \Big[\omega^{z_L}_d  \mathcal{T}(\beta_d | 0, \frac{\beta^\gamma}{\alpha^\gamma}, 2\alpha^\gamma) \nonumber \\ 
\label{eq:Z_approx}
&+ (1 - \omega^{z_L}_d)\mathcal{N}(\beta_d | m^{z_L}_d, \frac{\beta^\gamma}{\alpha^\gamma - 1} + v^{z_L})\Big]
\end{align}

@@

[ADD ON EP UPDATE OF PRIOR INCORPORATED]

\subsection{Hyperparameter optimisation}
The only hyperparameter in the proposed Bayesian \textsc{lista} is the shrinkage parameter $\lambda$. It can be optimised using the Type II maximum likelihood procedure. The Type II likelihood, i.e. the evidence $p(\boldsymbol\beta | \mathbf{y}, \lambda)$, of the Bayesian \textsc{lista} is equal to the normalisation constant $Z$ (\ref{eq:Z}) computed for the whole training dataset $\boldsymbol\beta$. Given the approximation~(\ref{eq:Z_approx}) the optimal hyperparameter $\lambda$ can be found by a gradient-based optimiser.

\section{Experiments}
\label{sec:experiments}
The proposed Bayesian \textsc{lista} is evaluated in the context of the sparse inverse problem, where the number of measurements $K$ is much smaller than the dimensionality of the vector $\boldsymbol\beta$.

@@

\subsection{Predictive performance on synthetic data}
First, the predictive performance of the proposed Bayesian \textsc{lista} is analysed on synthetic data. We generate $N_\text{train}=500$ sparse coefficients vectors $\boldsymbol\beta_n$ each of size $D = 100$. Coefficients $\boldsymbol\beta$ are generated from the spike-and-slab distribution with truncated slab: each component of $\beta_{nd}$ is zero with probability $0.8$ or is from standard Gaussian distribution without interval $(-0.1, 0.1)$ with probability $0.2$. Truncation of the zeros allows to remove from support pattern values with insignificant contributions \citep{xin2016maximal}. To simulate sparse observations, we have generated the random Gaussian design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$.  The observations are generated according to the \ref{eq:regression_problem} with zero-mean Gaussian noise with standard deviation $0.5$. $\lambda$ is assumed fixed as $0.1$.

@@

\subsection{Predictive performance on \textsc{mnist} data}
Here we evaluate the proposed Bayesian \textsc{lista} in terms of predictive performance on the \textsc{mnist} dataset \citep{lecun1998gradient}. The dataset contains images of handwritten digits of size $28 \times 28 = 784$. The design matrix $\mathbf{X}$ is learned on 5000 images with the minibatch online algorithm \citep{mairal2009online}. The resulting size of $\mathbf{X}$ is $K \times 784$. Then we generate observations as $\mathbf{y} = \mathbf{X}\boldsymbol\beta$, where $\boldsymbol\beta$ are images converted to vectors. We use randomly selected $100$ images for training and $100$ for validation. $\lambda$ is assumed fixed as $0.1$.

@@

\subsection{Active learning on \textsc{mnist}}
To demonstrate the potential scenario that can benefit from uncertainty estimates of the Bayesian \textsc{lista} we consider the active learning example \citep{settles.tr09}. Active learning area researches ways to select new training subsets to reduce total number of required supervision. One of the approaches for active learning is uncertainty sampling when the data with the least certain predictions is used for training. The uncertainty can be measured with entropy or variance. In our example no closed form for entropy of spike and slab exist, therefore we use variance from lemma \ref{thm:moments_spsl} as a measure of uncertainty.

@@

\section{Discussion}
\label{sec:discussion}
As far as the authors aware, this is the first implementation of Bayesian deep sparse coding algorithm. Although there are works on Bayesian sparsity in context of neural networks \citep{he2017bayesian}, they are not the Bayesian neural networks in the same sense as Bayesian \textsc{lista} but rather the interpretation of the sparse Bayesian  learning algorithm as the long short-term memory network. We find not only correct predictions but also useful posterior estimates for the predictions that show how the model is confident in its decisions.

>>>>

\section{\uppercase{Introduction}}
Though the idea of Bayesian learning in neural networks is not new \citep{neal2012bayesian}, it has regained its attention recently with the development of distributed approximate inference techniques \citep{li2015stochastic, hoffman2013stochastic}  and general boost in popularity of deep learning. In some spheres, such as self-driving cars or healthcare, uncertainty estimates are very important and Bayesian learning allows to obtain them.

@@

\section{\uppercase{Neural networks for sparse coding}}
\label{sec:nn_sc}
The sparse coding and sparse inverse problems are the linear regression with the assumption of sparse weights
\begin{equation}
\label{eq:regression_problem}
\mathbf{y} = \mathbf{X}\boldsymbol\beta + \varepsilon
\end{equation}
where $\mathbf{y} \in \mathbb{R}^K$ is the observation, $\mathbf{X} \in \mathbb{R}^{K \times D}$ is the design matrix, $\boldsymbol\beta \in \mathbb{R}^D$ is the unknown vector of weight with many elements $\beta_d$ equal to zero. $\varepsilon \in \mathbb{R}^K$ is the independent Gaussian noise.

@@

\section{\uppercase{Bayesian neural network for sparse coding}}
\label{sec:bayesian_lista}

This section presents the proposed Bayesian neural network for sparse coding. To formulate the Bayesian version of \textsc{lista} we impose the prior distributions on the unknown weights
\begin{subequations}
\label{eq:ws}
\begin{align}
&p(\mathbf{W}) = \prod_{d=1}^D\prod_{k=1}^K \mathcal{N}(w_{ij} | 0, \eta^{-1}), \\
&p(\mathbf{S}) = \prod_{d'=1}^D\prod_{d''=1}^D \mathcal{N}(s_{d'd''} | 0, \eta^{-1}),
\end{align}
\end{subequations}
where $w_{ij}$ is a component of the matrix $\mathbf{W}$, $s_{d'd''}$ is a component of the matrix $\mathbf{S}$, $\eta$ is the precision of the Gaussian distribution.
The likelihood of $\mathbf{B}$ is defined as 
\begin{align}
p(&\mathbf{B}| \mathbf{Y}, \mathbf{W}, \mathbf{S}, \gamma, \lambda) \nonumber \\
\label{eq:likelihood}
= &\prod_{n=1}^N\prod_{d=1}^D\mathcal{N}(\beta_d^{(n)}; [f(\mathbf{y^{(n)}}; \mathbf{W}, \mathbf{S}, \lambda)]_d, \gamma^{-1}),
\end{align}
where $[\cdot]_d$ is the d-th component of a vector, $f(\mathbf{y}^{(n)}; \mathbf{W}, \mathbf{S}, \lambda)$ is the output of the Bayesian \textsc{lista} neural network, $\gamma$ is the precision of the Gaussian distribution.
The prior of the introduced Gaussian precisions are set to the Gamma distribution with parameters $a^{\cdot}$ and $b^{\cdot}$:
\begin{subequations}
\label{eq:gamma_eta}
\begin{align}
&p(\gamma) = \text{Gam}\left(\gamma; a^{\gamma}, b^{\gamma}\right), \\
&p(\eta) = \text{Gam}\left(\eta; 	a^{\eta}, b^{\eta}\right)
\end{align}
\end{subequations}

@@

\section{\uppercase{Uncertainty propagation through soft thresholding}}
\label{sec:fprop}
For every layer of the Baysian \textsc{lista} we assume that $\widehat{\boldsymbol\beta}_{l-1}$ has the spike and slab distribution with parameters: $\boldsymbol\omega$ is a probability of a spike, $\mathbf{m}$ is a mean of a slab Gaussian distribution, and $\mathbf{v}$ is a variance of the slab distribution:
\begin{equation}
[\widehat{\boldsymbol\beta}_{l-1}]_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d),
\end{equation}
where $\omega_d$, $m_d$, and $v_d$ are components of vectors $\boldsymbol\omega$, $\mathbf{m}$, and $\mathbf{v}$, respectively, $\delta_0$ is the delta-function that represents a spike. 
We initialise $\widehat{\boldsymbol\beta}_{0}$ with the spike and slab distribution with parameters $\boldsymbol\omega = \mathbf{0}$, $\mathbf{m} = \mathbf{0}$, and $\mathbf{v} = \mathbf{1}$.

@@

  \begin{lemma}[Product of Gaussian matrix and spike and slab vector]
  \label{thm:matrix_vector}
 Let $\mathbf{S} \in \mathbb{R}^{D \times D}$ be a matrix of independent Gaussian-distributed random variables: $s_{d'd''} \sim \mathcal{N}(m^s_{d'd''}, v^s_{d'd''})$, and $\widehat{\boldsymbol\beta }\in \mathbb{R}^D$ be a vector with spike-and-slab distributed variables: $\widehat{\beta}_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d)$. If their product is approximated as a vector $\mathbf{e} \in \mathbb{R}^{D}$ of random variables $e_d \sim \mathcal{N}(m^e_d, w^e_d)$, then 
\begin{subequations}
\begin{align}
 m^e_d &= \sum_{d'=1}^D m^s_{dd'}(1-\omega_{d'})m_{d'}, \\
 \begin{split}
 v^e_d &= \sum_{d'=1}^D [(m^s_{dd'})^2(1-\omega_{d'})^2v_{d'} \\
 & {}+ (1-\omega_{d'})^2(m_{d'})^2v^s_{dd'} + v^s_{dd'}(1-\omega_{d'})^2v_{d'}].
 \end{split}
 \end{align}
\end{subequations}
 \end{lemma}
 \begin{proof}
 	We compute the mean and variance of the product $\mathbf{S}\widehat{\boldsymbol\beta}$ with an assumption of independence of its components and approximate resulting distribution with Gaussian. The quality of this approximation is discussed in section \ref{sec:approx_quality}
\begin{flalign*}
	\mathbb{E}e_d &= \sum_{d'=1}^D \mathbb{E}[s_{dd'}\widehat{\beta}_{d'}]  = \sum_{d'=1}^D m^s_{dd'}\mathbb{E}\widehat{\beta}_{d'}\\
	\operatorname{Var}e_d &= \sum_{d'=1}^D \operatorname{Var}[s_{dd'}\widehat{\beta}_{d'}] = \sum_{d'=1}^D [(\mathbb{E}s_{dd'})^2 \operatorname{Var}\widehat{\beta}_{d'} \\
	&{}+ (\mathbb{E}\widehat{\beta}_{d'})^2 \operatorname{Var}s_{dd'} + \operatorname{Var}\widehat{\beta}_{d'} \operatorname{Var}s_{dd'}]
\end{flalign*}
where $\mathbb{E}\widehat{\beta}_{d'}$, $\operatorname{Var}\widehat{\beta}_{d'}$ are computed according to lemma~\ref{thm:moments_spsl}.
 \end{proof}

@@

The proposed operations on Gaussian and spike-and-slab distributions allow to formulate the main result in this section that describes how the uncertainty propagation works.
\begin{theorem}[Bayesian \textsc{lista} forward propagation]
\label{thm:prob_layer}
The spike and slab distribution for $\widehat{\boldsymbol\beta}$ with can be propagated through the \textsc{lista} layer and the parameters after propagation can be computed as following:
\begin{enumerate}
	\item $\mathbf{b} = \mathbf{W}\mathbf{y}$ is computed according to lemma \ref{thm:matrix_const};
	\item $\mathbf{e} = \mathbf{S}\widehat{\boldsymbol\beta}_{l-1}$ is computed according to lemma \ref{thm:matrix_vector};
	\item $\mathbf{c} = \mathbf{b} + \mathbf{e}$ is computed according to lemma \ref{thm:sum_vectors};
	\item $\widehat{\boldsymbol\beta}_{l} = h_\lambda(\mathbf{c})$ is computed according to lemma \ref{thm:soft_thresholding}.
\end{enumerate}
\end{theorem}

\subsection{\uppercase{Approximation quality}}
\label{sec:approx_quality}
We have used two approximations in forward propagation of uncertainty. First, in lemma \ref{thm:matrix_vector} a Gaussian matrix is multiplied by a spike-and-slab vector and their product is approximated with the Gaussian distribution. Second, in lemma \ref{thm:soft_thresholding} the result of soft thresholding of a Gaussian vector is approximated with the spike-and-slab distribution. In this section we demonstrate that these approximations are close to the real distributions.

@@

\section{\uppercase{Backpropagation}}
\label{sec:backpropagation}

The exact posterior (\ref{eq:posterior}) is approximated with a factorised distribution
\begin{align}
\label{eq:approximating_dsitribution}
\begin{split}
q(\mathbf{W}, \mathbf{S}, \gamma, \eta) &= \prod_{d=1}^D\prod_{k=1}^K \mathcal{N}(w_{dk} | m^w_{dk}, v^w_{dk}) \\
&\times \prod_{d'=1}^D\prod_{d''=1}^D \mathcal{N}(s_{d'd''} | m^s_{d'd''}, v^s_{d'd''}) \\
&\times \text{Gam}(\gamma; a^\gamma, b^\gamma) \text{Gam}(\eta; a^\eta, b^\eta) 
\end{split}
\end{align}

For Bayesian inference we expand the probabilistic backpropagation algorithm~\citep{hernandez2015probabilistic} for computing parameter updates. It is based on assumed density filtering (ADF) and expectation propagation (EP) and allows to update parameters of the distributions based on the derivative of the logarithm of a normalisation constant. ADF is based on iterative incorporation of factors from the true posterior $p$ (\ref{eq:posterior}) into the factorised approximating distribution $q$ (\ref{eq:approximating_dsitribution}), whereas in EP factors in the $q$ are iteratively replaced by factors from $p$.

When a factor $p$ is incorporated into $q$, $q$ as a function of Gaussian-distributed weights $W$  and $S$ has the following form:
\begin{equation}
q(\cdot) = Z^{-1}f(\cdot)\mathcal{N}(\cdot; m, v)
\end{equation}
where $Z$ is the normalisation constant and $f(\cdot)$ - arbitrary function.

According to \cite{minka2001thesis}, new parameters of Gaussian distribution for$ \cdot$ can be computed as
\begin{align}
m^{\text{new}} &= m + v \frac{\partial \log Z}{\partial m}, \\
v^{\text{new}} &= v - v^2\left[ \left(\frac{\partial \log Z}{\partial m}\right)^2 - 2 \frac{\partial \log Z}{\partial v}\right]
\end{align}

Then to find new values of $W$, $S$ we need to compute the $Z$ every time when the factor of the true posterior is incorporated in $q$.

\subsection{\uppercase{Likelihood}}

With the likelihood factors~(\ref{eq:likelihood}) of the true posterior we employ the ADF approach and iteratively incorporate them into the approximating distribution $q$. The normalisation constant of the approximating distribution $q$ with the likelihood term for the data point $n$ incorporated can be computed as follows (to simplify notation the superscript $(n)$ is omitted)
\begin{align}
\label{eq:Z}
\begin{split}
Z = \int \prod_{d=1}^{D} \big[&\mathcal{N}(\beta_d | f(\mathbf{y} ; \mathbf{S}, \mathbf{W}, \lambda), \gamma^{-1}) \\
 &q(\mathbf{W}, \mathbf{S}, \gamma, \eta)\big] \mathrm{d}\mathbf{W} \mathrm{d}\mathbf{S} \mathrm{d}\gamma \mathrm{d}\eta
 \end{split}
\end{align}
We sample $\mathbf{W}$, $\mathbf{S}$ from $q$ and get $\widehat{\boldsymbol\beta}_L = f(\mathbf{y} ; \mathbf{S}, \mathbf{W}, \lambda)$, that is the output from the network and approximated with the spike and slab distribution with parameters $\boldsymbol\omega^{\widehat{\boldsymbol\beta}_L}$, $\mathbf{m}^{\widehat{\boldsymbol\beta}_L}$, and $\mathbf{v}^{\widehat{\boldsymbol\beta}_L}$
\begin{align}
&Z \approx \int \prod_{d=1}^{D} \big[\mathcal{N}(\beta_d | [\widehat{\boldsymbol\beta}_L]_d, \gamma^{-1}) \nonumber \\
&\times (\omega^{\widehat{\boldsymbol\beta}_L}_d \delta_0([\widehat{\boldsymbol\beta}_L]_d) + (1 - \omega^{\widehat{\boldsymbol\beta}_L}_d)\mathcal{N}([\widehat{\boldsymbol\beta}_L]_d | m^{\widehat{\boldsymbol\beta}_L}_d, v^{\widehat{\boldsymbol\beta}_L}_d)) \nonumber\\
&\times \text{Gam} (\gamma_d; \alpha^\gamma, \beta^\gamma)\big]\mathrm{d}\widehat{\boldsymbol\beta}_L \mathrm{d}\gamma  \nonumber\\
&= \prod_{d=1}^{D} \Big[\omega^{\widehat{\boldsymbol\beta}_L}_d \int \big[\mathcal{N}(\beta_d | [\widehat{\boldsymbol\beta}_L]_d, \gamma^{-1})  \delta_0([\widehat{\boldsymbol\beta}_L]_d) \nonumber \\
&\times \text{Gam} (\gamma; \alpha^\gamma, \beta^\gamma)\big]\mathrm{d}[\widehat{\boldsymbol\beta}_L]_d \mathrm{d}\gamma  \nonumber\\
& + (1 - \omega^{\widehat{\boldsymbol\beta}_L}_d)\int \big[\mathcal{N}(\beta_d | [\widehat{\boldsymbol\beta}_L]_d, \gamma^{-1})\nonumber\\
&\times \mathcal{N}([\widehat{\boldsymbol\beta}_L]_d | m^{\widehat{\boldsymbol\beta}_L}_d, v^{\widehat{\boldsymbol\beta}_L}_d)) \text{Gam} (\gamma; \alpha^\gamma, \beta^\gamma)\big]\mathrm{d}[\widehat{\boldsymbol\beta}_L]_d \mathrm{d}\gamma\Big]  \nonumber\\
& = \prod_{d=1}^{D} \Big[\omega^{\widehat{\boldsymbol\beta}_L}_d \int \mathcal{N}(\beta_d | 0, \gamma^{-1})  \text{Gam} (\gamma; \alpha^\gamma, \beta^\gamma) d\gamma  \nonumber\\
& + (1 - \omega^{\widehat{\boldsymbol\beta}_L}_d)\int \big[\mathcal{T}(\beta_d | [\widehat{\boldsymbol\beta}_L]_d, \frac{\beta^\gamma}{\alpha^\gamma}, 2\alpha^\gamma)  \nonumber\\
&\times \mathcal{N}([\mathbf{\widehat{\boldsymbol\beta}_L}]_d| m^{\widehat{\boldsymbol\beta}_L}_d, v^{\widehat{\boldsymbol\beta}_L}_d))\big] \mathrm{d}[\widehat{\boldsymbol\beta}_L]_d\Big]  \nonumber\\
& \approx \prod_{d=1}^D \Big[\omega^{\widehat{\boldsymbol\beta}_L}_d  \mathcal{T}(\beta_d | 0, \frac{\beta^\gamma}{\alpha^\gamma}, 2\alpha^\gamma) \nonumber \\ 
\label{eq:Z_approx}
&+ (1 - \omega^{\widehat{\boldsymbol\beta}_L}_d)\mathcal{N}(\beta_d | m^{\widehat{\boldsymbol\beta}_L}_d, \frac{\beta^\gamma}{\alpha^\gamma - 1} + v^{\widehat{\boldsymbol\beta}_L})\Big]
\end{align}

@@

\subsection{\uppercase{Prior}}
Prior factors from $p$ (\ref{eq:ws}, \ref{eq:gamma_eta}) are incorporated into $q$ with the EP algorithm, i.e. they replace the corresponding approximating factors from $q$ and then $q$ is updated to minimise its KL divergence with $q$ that has the true factor incorporated.

\subsection{\uppercase{Hyperparameter optimisation}}
The only hyperparameter in the proposed Bayesian \textsc{lista} is the shrinkage parameter $\lambda$. It can be optimised using the Type II maximum likelihood procedure. The Type II likelihood, i.e. the evidence $p(\boldsymbol\beta | \mathbf{y}, \lambda)$, of the Bayesian \textsc{lista} is equal to the normalisation constant $Z$ (\ref{eq:Z}) computed for the whole training dataset $\boldsymbol\beta$. Given the approximation~(\ref{eq:Z_approx}) the optimal hyperparameter $\lambda$ can be found by a gradient-based optimiser.

\section{\uppercase{Experiments}}
\label{sec:experiments}
The proposed Bayesian \textsc{lista} is evaluated in the context of the sparse inverse problem, where the number of measurements $K$ is much smaller than the dimensionality of the vector $\boldsymbol\beta$.

@@

\subsection{\uppercase{Predictive performance on synthetic data}}
First, the predictive performance of the proposed Bayesian \textsc{lista} is analysed on synthetic data. We generate $N_\text{train}=500$ sparse coefficients vectors $\boldsymbol\beta_n$ each of size $D = 100$. Coefficients $\boldsymbol\beta$ are generated from the spike-and-slab distribution with truncated slab: each component of $\beta_{nd}$ is zero with probability $0.8$ or is from standard Gaussian distribution without interval $(-0.1, 0.1)$ with probability $0.2$. Truncation of the zeros allows to remove from support pattern values with insignificant contributions \citep{xin2016maximal}. To simulate sparse observations, we have generated the random Gaussian design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$.  The observations are generated according to the \ref{eq:regression_problem} with zero-mean Gaussian noise with standard deviation $0.5$. $\lambda$ is assumed fixed as $0.1$.

@@

\subsection{\uppercase{Predictive performance on mnist data}}
Here we evaluate the proposed Bayesian \textsc{lista} in terms of predictive performance on the \textsc{mnist} dataset \citep{lecun1998gradient}. The dataset contains images of handwritten digits of size $28 \times 28 = 784$. The design matrix $\mathbf{X}$ is learned on 5000 images with the minibatch online algorithm \citep{mairal2009online}. The resulting size of $\mathbf{X}$ is $K \times 784$. Then we generate observations as $\mathbf{y} = \mathbf{X}\boldsymbol\beta$, where $\boldsymbol\beta$ are images converted to vectors. We use randomly selected $100$ images for training and $100$ for validation. $\lambda$ is assumed fixed as $0.1$.

@@

\subsection{\uppercase{Active learning on mnist}}
To demonstrate the potential scenario that can benefit from uncertainty estimates of the Bayesian \textsc{lista} we consider the active learning example \citep{settles.tr09}. Active learning area researches ways to select new training subsets to reduce total number of required supervision. One of the approaches for active learning is uncertainty sampling when the data with the least certain predictions is used for training. The uncertainty can be measured with entropy or variance. In our example no closed form for entropy of spike and slab exist, therefore we use variance from lemma \ref{thm:moments_spsl} as a measure of uncertainty.

@@

\section{\uppercase{Discussion}}
\label{sec:discussion}
As far as the authors aware, this is the first implementation of Bayesian deep sparse coding algorithm. Although there are works on Bayesian sparsity in context of neural networks \citep{he2017bayesian}, they are not the Bayesian neural networks in the same sense as Bayesian \textsc{lista} but rather the interpretation of the sparse Bayesian  learning algorithm as the long short-term memory network. We find not only correct predictions but also useful posterior estimates for the predictions that show how the model is confident in its decisions.
